Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-07-26 ms_reorg.tex
adding optical depth information to steady state spec plot

The main impediment to performing magnetohydrodynamic simulations with fully a coupled radiative transfer solution via Monte Carlo methods is the prohibitive computational cost. A primary factor in the cost of Monte Carlo is that the number of scatterings a photon undergoes is proportional to the line center optical depth.  Near the base of the atmosphere, individual simulation cells can have optical depths of more than a million, so most of the time is spent integrating photons in these cells. This means that any method that can treat these zone via an analytic solution has the potential to greatly accelerate the calculation.

\section{ previous analytic work }

@@

\subsection{Derivation of the Steady-State Solution}
Now, consider a sphere of radius $R$ and uniform density, with line-center optical depth $\tau_0 = (k/\sqrt{\pi}\Delta)R$. After changing variables from $\nu$ to $\sigma$ using $d\sigma = \sqrt{2/3}d\nu/(\phi \Delta^2)$ in Eq. \ref{eq:rteqn} and performing all angular integrals, we obtain Eq. \ref{app:rteqn_derivation}\ref{eq:finaleqn} (see Appendix \ref{app:rteqn_derivation}) and set $p=\alpha_{\rm abs}=0$ to eliminate photon destruction terms. With a photon emission term given by Eq. \ref{app:rteqn_derivation}\ref{eq:jem}, this gives the transfer equation
\be
\nabla^2 J + \left( \frac{k}{\Delta} \right)^2 \frac{\partial^2 J}{\partial \sigma^2} & = & 
- \frac{ \sqrt{6} kL}{4\pi \Delta^2} \delta^3(\vec{x} - \vec{x}_s) \delta (\sigma - \sigma_{\rm s})
\label{eq:rt_no_destr}
\ee
where the form of the source term is in Eq. \ref{app:rteqn_derivation}\ref{eq:jem_v2}. The boundary condition is that there is no incoming flux at the surface, which can be written
\citep{1986rpa..book.....R}
\be
J & = & \sqrt{3} H
\label{eq:bc}
\ee
at $r=R$. An analytic solution $J_d$ which is divergent at the center of the sphere and at the emission frequency may be found. However, it is not a good approximation to the true solution, as it is too large at $r=R$ by a factor of $J_{\rm d}(R,\sigma)/ H_{\rm d}(R,\sigma) \sim a\tau_0/x^2 \sim (a\tau_0)^{1/3} \gg 1$. This solution is included in Fig. \ref{fig:sol_mc_residual_0} for qualitative understanding, but will not be discussed further.

@@

\subsection{Scaling with Line Center Optical Depth}
To examine the relative size of the corrective term $H_{\rm bc}$ as compared to $H_0$, we consider the case where the optical depth is large and thus the mean intensity will be considerably larger than the surface flux. Starting from Eq. \ref{eq:bc2}, we assume $H_{\rm bc} \ll J_{\rm bc}$ such that
\be
J_{\rm bc}(R,\sigma) & \approx & \sqrt{3} H_0(R,\sigma).
\label{eq:bc4}
\ee
The Fourier transform for $J_{\rm bc}$ at the surface becomes
\be
J_{\rm bc}(R, \sigma) \equiv \int_{-\infty}^{\infty} \frac{ds}{2\pi} e^{is\sigma} A(s)
\ee
with Fourier amplitudes
\be \label{eq:amps}
A(s) = \sqrt{3} \int_{-\infty}^{\infty} d\sigma e^{-is\sigma} H_0(R, \sigma).
\ee
Plugging the expression for $H_0$ in Eq. \ref{eq:H0surf} into Eq. \ref{eq:amps} yields an expression for the Fourier amplitudes
\be \label{eq:plugged_amps}
A(s) = \sqrt{3} \int_{-\infty}^\infty d\sigma e^{-is\sigma}\left(\frac{1}{3k\phi(\sigma)}\frac{\sqrt{6}}{32\pi}\frac{L}{\Delta R^3}\frac{1}{1+\cosh \left[\frac{\pi \Delta}{kR} |\sigma - \sigma_i|\right]}\right)
\ee
with
\be
\phi(\sigma) = \left(\frac{2a}{27\pi}\right)^{1/3}\frac{1}{\Delta |\sigma|^{2/3}}.
\ee
With the line center optical depth defined as
\be \label{eq:tau0}
\tau_0 = \frac{kR}{\sqrt{\pi}\Delta}
\ee
we scale the variables $\sigma$ and $s$ by $\tau_0$, now using characteristic photon frequency $\bar \sigma = \sigma / \tau_0$ and characteristic wavenumber $\bar s = s \tau_0$. Eq. \ref{eq:plugged_amps} becomes
\be
A(\bar s) = \frac{\sqrt{2}}{32\pi}\left(\frac{27\pi}{2}\right)^{1/3}\frac{L}{kR^3}\frac{\tau_0^{5/3}}{a^{1/3}}\int_{-\infty}^\infty d\bar\sigma e^{-i\bar s \bar\sigma} |\bar\sigma^{2/3}|\frac{1}{1+\cosh \left[\sqrt{\pi}|\bar\sigma - \bar\sigma_i|\right]}.
\ee
These coefficients are used to evaluate the boundary condition's contribution to the flux (Eq. \ref{eq:Hbc}) at the surface of the sphere. Dividing this additional flux $H_{\rm bc}$ by the contribution from the $J=0$ solution, $H_0$ (Eq. \ref{eq:H0surf}), yields an overall proportionality
\be \label{eq:hbc_scaling}
\frac{H_{\rm bc}(R, \sigma)}{H_0(R, \sigma)} \propto \frac{1}{(a\tau_0)^{1/3}}.
\ee
Thus, the fractional importance of the corrective frequency-dependent boundary condition term scales as the inverse cube root of the damping parameter $a$ times the line-center optical depth $\tau_0$. At very large optical depths, it is expected that the corrective term will be small, but will become increasingly important in less optically thick media.

@@

    \caption{The scaling of numerically-obtained resonant frequencies as compared with the analytic expression in Eq. \ref{eq:gamma_nm}.}

@@

The values of the $\gamma_{nm}$ resonant frequencies are calculable analytically. Their scale depends on $m$, $n$, and the factor $(a\tau_0)^{1/3}$ according to 

@@

\begin{figure}
    \centering
    \includegraphics{fitting_formula.pdf}
    \caption{A wait time distribution of escaping photons for two different line center optical depths $\tau_0$, which is fit using Eq. \ref{eq:fitting_function}. A sum over 20 spatial eigenmodes and 500 frequency eigenmodes is labeled ``Eigenfunctions''. All calculations were performed with a monochromatic source of photons at line center ($x_s = 0.0$).}
    \label{fig:escape_time}
\end{figure}

@@

>>>>

The main impediment to performing magnetohydrodynamic simulations with fully a coupled radiative transfer solution via Monte Carlo methods is the prohibitive computational cost. A primary factor in the cost of Monte Carlo is that the number of scatterings a photon undergoes is proportional to the line center optical depth.  Near the base of the atmosphere, individual simulation cells can have optical depths of more than a million, so most of the time is spent integrating photons in these cells. This means that any method that can treat these zone via an analytic solution has the potential to greatly accelerate the calculation. A complete, general analytic solution for the resonance scattering has never been derived but approximate solutions exist in certain limits.

{\color{red} Moved text on Harrington into the introduction since it is a primary driver of the calculation.}

@@

\subsection{Derivation of the Steady-State Solution}

{\color{red} Moved the time dependent part back to the appendix. It is probably sufficient to just state eq. \ref{eq:finaleqn} as the starting point, referring to the reader to the appendix. We may want a short paragraph defining some of the key notation, though, so the reader is not forced to extensively reference the appendix.}

Now, consider a sphere of radius $R$ and uniform density, with line-center optical depth $\tau_0 = (k/\sqrt{\pi}\Delta)R$. After changing variables from $\nu$ to $\sigma$ using $d\sigma = \sqrt{2/3}d\nu/(\phi \Delta^2)$ in Eq. \ref{eq:rteqn} and performing all angular integrals, we obtain Eq. \ref{app:rteqn_derivation}\ref{eq:finaleqn} (see Appendix \ref{app:rteqn_derivation}) and set $p=\alpha_{\rm abs}=0$ to eliminate photon destruction terms. With a photon emission term given by Eq. \ref{app:rteqn_derivation}\ref{eq:jem}, this gives the transfer equation
\be
\nabla^2 J + \left( \frac{k}{\Delta} \right)^2 \frac{\partial^2 J}{\partial \sigma^2} & = & 
- \frac{ \sqrt{6} kL}{4\pi \Delta^2} \delta^3(\vec{x} - \vec{x}_s) \delta (\sigma - \sigma_{\rm s})
\label{eq:rt_no_destr}
\ee
where the form of the source term is in Eq. \ref{app:rteqn_derivation}\ref{eq:jem_v2}. The boundary condition is that there is no incoming flux at the surface, which can be written
\citep{1986rpa..book.....R}
\be
J & = & \sqrt{3} H
\label{eq:bc}
\ee
at $r=R$. An analytic solution $J_d$ which is divergent at the center of the sphere and at the emission frequency may be found. However, it is not a good approximation to the true solution, as it is too large at $r=R$ by a factor of $J_{\rm d}(R,\sigma)/ H_{\rm d}(R,\sigma) \sim a\tau_0/x^2 \sim (a\tau_0)^{1/3} \gg 1$. This solution is included in Fig. \ref{fig:sol_mc_residual_0} for qualitative understanding, but will not be discussed further.

{\color{red} We do not need an extensive discussion fo the divergent solution, but we should probably right it down here for compelteness, since we include it in the figure.}

@@

\subsection{Scaling with Line Center Optical Depth}

{\color{red} This section can probably be shortened to a couple of paragraphs with most of the equations removed and the results just stated.}

To examine the relative size of the corrective term $H_{\rm bc}$ as compared to $H_0$, we consider the case where the optical depth is large and thus the mean intensity will be considerably larger than the surface flux. Starting from Eq. \ref{eq:bc2}, we assume $H_{\rm bc} \ll J_{\rm bc}$ such that
\be
J_{\rm bc}(R,\sigma) & \approx & \sqrt{3} H_0(R,\sigma).
\label{eq:bc4}
\ee
The Fourier transform for $J_{\rm bc}$ at the surface becomes
\be
J_{\rm bc}(R, \sigma) \equiv \int_{-\infty}^{\infty} \frac{ds}{2\pi} e^{is\sigma} A(s)
\ee
with Fourier amplitudes
\be \label{eq:amps}
A(s) = \sqrt{3} \int_{-\infty}^{\infty} d\sigma e^{-is\sigma} H_0(R, \sigma).
\ee
Plugging the expression for $H_0$ in Eq. \ref{eq:H0surf} into Eq. \ref{eq:amps} yields an expression for the Fourier amplitudes
\be \label{eq:plugged_amps}
A(s) = \sqrt{3} \int_{-\infty}^\infty d\sigma e^{-is\sigma}\left(\frac{1}{3k\phi(\sigma)}\frac{\sqrt{6}}{32\pi}\frac{L}{\Delta R^3}\frac{1}{1+\cosh \left[\frac{\pi \Delta}{kR} |\sigma - \sigma_i|\right]}\right)
\ee
with
\be
\phi(\sigma) = \left(\frac{2a}{27\pi}\right)^{1/3}\frac{1}{\Delta |\sigma|^{2/3}}.
\ee
With the line center optical depth defined as
\be \label{eq:tau0}
\tau_0 = \frac{kR}{\sqrt{\pi}\Delta}
\ee
we scale the variables $\sigma$ and $s$ by $\tau_0$, now using characteristic photon frequency $\bar \sigma = \sigma / \tau_0$ and characteristic wavenumber $\bar s = s \tau_0$. Eq. \ref{eq:plugged_amps} becomes
\be
A(\bar s) = \frac{\sqrt{2}}{32\pi}\left(\frac{27\pi}{2}\right)^{1/3}\frac{L}{kR^3}\frac{\tau_0^{5/3}}{a^{1/3}}\int_{-\infty}^\infty d\bar\sigma e^{-i\bar s \bar\sigma} |\bar\sigma^{2/3}|\frac{1}{1+\cosh \left[\sqrt{\pi}|\bar\sigma - \bar\sigma_i|\right]}.
\ee
These coefficients are used to evaluate the boundary condition's contribution to the flux (Eq. \ref{eq:Hbc}) at the surface of the sphere. Dividing this additional flux $H_{\rm bc}$ by the contribution from the $J=0$ solution, $H_0$ (Eq. \ref{eq:H0surf}), yields an overall proportionality
\be \label{eq:hbc_scaling}
\frac{H_{\rm bc}(R, \sigma)}{H_0(R, \sigma)} \propto \frac{1}{(a\tau_0)^{1/3}}.
\ee
Thus, the fractional importance of the corrective frequency-dependent boundary condition term scales as the inverse cube root of the damping parameter $a$ times the line-center optical depth $\tau_0$. At very large optical depths, it is expected that the corrective term will be small, but will become increasingly important in less optically thick media.

@@

    \caption{The scaling of numerically-obtained resonant frequencies as compared with the analytic expression in Eq. \ref{eq:gamma_nm}. The data shown is for $\tau_0=10^7$ and $x_s=0.0$.}

@@

The values of the $\gamma_{nm}$ resonant frequencies can be described approximately with Eq. \ref{eq:gamma_nm}. Their scale depends on $m$, $n$, and the factor $(a\tau_0)^{1/3}$ according to 

@@

\begin{figure}
    \centering
    \includegraphics{waittime.pdf}
    \caption{A wait time distribution of escaping photons for two different line center optical depths $\tau_0$, which is fit using Eq. \ref{eq:fitting_function}. A sum over 20 spatial eigenmodes and 500 frequency eigenmodes is labeled ``Eigenfunctions''. All calculations were performed with a monochromatic source of photons at line center ($x_s = 0.0$).}
    \label{fig:escape_time}
\end{figure}

@@

\section{ Discussion }

{\color{red} Moved the acceleration content to discussion}
